Evaluating infectious disease forecasts with allocation scoring rules

### 2023-05-01 alloscore_manuscript/allocation.tex
update section on allocation score to have better flow after new quantile score section

Suppose that a decision-maker is tasked with determining the quantity $x$ of a protective measure to procure; for example, $x$ might represent the number of hospital beds or amount of medicine to purchase. Additionally, suppose that each unit of this good has cost $C$ so that the total cost of procurement is $Cx$. The variable $y$ denotes the eventual realized need for this resource, e.g. the number of patients in need of a hospital bed or the amount of medication that is needed. We assume that each unit of unmet need incurs a loss denoted by $L$, so that if the selected procurement level $x$ is less than the realized need $y$, a loss of $L(x-y)$ results. With this formalization of a decision-making task, we can proceed to develop a proper scoring rule using the procedure outlined in section \ref{sec:methods.decisiontheory}.

@@

We now develop a scoring rule for probabilistic forecasts that measures the value of a forecast as an input to decision making about how to allocate limited resources to meet demand across multiple locations. As a concrete example, we take the resource to be a good such as ventilators or oxygen supply. An administrator is tasked with determining where to send these resources so as to meet demand among hospital patients in different facilities or states.

We first specialize our notation to this decision-making setting. We define an action $\mathbf{x} = (x_1, \ldots, x_n)$ as a vector specifying the amount that is allocated to each of the $n$ locations. We require that each $x_i$ is non-negative, and that the total allocation across all locations does not exceed a constraint $K$ on the total available resources: $\sum_{i=1}^n x_i \leq K$. The set $\mathcal{X}$ collects all possible allocations that satisfy these constraints. The eventually realized resource demand in each location is denoted by $\mathbf{y} = (y_1, \ldots, y_n)$. At the time that a decision-maker sets the resource allocation, the demand $\mathbf{y}$ is not yet known. We therefore define the random vector $Y = (Y_1, \ldots, Y_n)$ where $Y_i$ represents the as-yet-unknown level of resource demand in location $i$. The forecast $F = (F_1, \ldots, F_n)$ collects forecasts of demand in each location. Here we identify $F_i$ with its cumulative distribution function (CDF), and $F_i^{-1}$ denotes the quantile function. We assume that the forecasts do not allow for the possibility of negative demand, i.e. the support of each $F_i$ is a subset of $\mathbb{R}^+$. With this notation in place, we can proceed to develop a proper scoring rule following the outline in the previous section.

\paragraph{Step 1: specify a loss function.} The loss associated with a particular allocation is calculated by comparing the amount allocated to each location to the realized resource demand in that location. Specifically, suppose that there is a marginal cost $L$ that accrues for each unit of demand that is not met. We can calculate the allocation loss across all locations as
\begin{equation}
s(\mathbf{x}, \mathbf{y}) = \sum_{i=1}^n L (x_i - y_i)_-, \label{eqn:loss_fn}
\end{equation}
where $(x_i - y_i)_- := \max(-(x_i - y_i),0)$ is $0$ if the amount $x_i$ allocated to unit $i$ is greater than or equal to the realized demand $y_i$ in that location; otherwise, it is $y_i - x_i$, the amount of unmet need in that location. We note that a number of generalizations to this loss specification have been formulated in the literature, including an allowance for costs for over-allocation to a particular unit (e.g. if there are storage costs for unused resources), differing losses different units (e.g. if a unit of unmet demand imposes more severe costs in one location than another), and the introduction of a convex function that controls the rate at which costs accrue depending on the scale of need. We consider these and other generalizations in other work, but for the present exposition we restrict our attention to the relatively simple loss formulation of Equation \eqref{eqn:loss_fn}.

\paragraph{Step 2: Given a probabilistic forecast $F$, identify the Bayes act.} The Bayes act is the allocation that minimizes the expected loss:
\begin{align}
    \mathbf{x}^F &= \underset{\mathbf{x}\in \mathbb{R}^N, 0 \leq \mathbf{x}}{\mathrm{arg min}}\,\, \sbar_F(\mathbf{x}) \text{ subject to } 
    \sum_{i=1}^N x_i \leq K \text{, where} \label{AP} \\
    \sbar_F(\mathbf{x}) &= \mathbb{E}_{F} s(\mathbf{x}, \mathbf{Y}) = \sum_{i=1}^{N} \mathbb{E}_{F_i}[s(x_i, Y_i)] \label{eqn:sbar}
\end{align}
It can be shown that with the loss function given in Equation \eqref{eqn:loss_fn}, the Bayes act has $x^F_i = F_i^{-1}(1 - \lambda^{\star}/L)$, where $\lambda^{\star}$ is chosen so as to satisfy the equation
\begin{align}
\sum_{i=1}^{N}F_i^{-1}(1 - \lambda^{\star}/L) = K.
\end{align}
This partial solution to the allocation problem seems to have first appeared in [@hadleywhitin1963]; see the supplemental materials for a derivation.

One interpretation of this result is that the Bayes act sets the allocation in each location $i$ to a quantile of the forecast distribution $F_i$ for that location. The quantile is at a probability level $(1 - \lambda^{\star}/L)$ that is the same for all locations, and is chosen such that the constraint is satisfied. An alternative interpretation comes from noting that for each location $i$, $\frac{\partial}{\partial x_i} \sbar_F(\mathbf{x}) \vert_{\mathbf{x}=\mathbf{x}^F} = \lambda^{\star}$ (see the supplement for a proof).
%$\frac{\partial}{\partial x_i} \sbar_F(\mathbf{x}) = L[F_i(x_i) - 1]$ (see the supplement for a derivation), and evaluating at the Bayes allocation we arrive at
%\begin{align}
%\frac{\partial}{\partial x_i} \sbar_F(\mathbf{x}) \vert_{\mathbf{x}=\mathbf{x}^F} &= L[F_i(x^F_i) - 1] = L[F_i(F_i^{-1}(1 - \lambda^{\star}/L)) - 1] = \lambda^{\star}.
%\end{align}
In words, at the allocation given by the Bayes act, the rate of change of the expected score as a function of the amount allocated to location $i$ is given by $\lambda^{\star}$.
This derivative is the same for all locations, so the optimal allocation divides the available resources across all locations in such a way that according to $F$, the expected benefit of 1 additional unit of resources is the same in all locations.

>>>>

Suppose that a decision-maker is tasked with determining the quantity $x$ of a protective measure to procure; for example, $x$ might represent the number of hospital beds or amount of medicine to purchase. Additionally, suppose that each unit of this good has cost $C$ so that the total cost of procurement is $Cx$. The variable $y$ denotes the eventual realized need for this resource, e.g. the number of patients in need of a hospital bed or the amount of medication that is needed. We assume that each unit of unmet need incurs a loss denoted by $L$, so that if the selected procurement level $x$ is less than the realized need $y$, a loss of $L(x-y)$ results. At the time that a decision-maker determines the amount $x$ to procure, the demand $y$ is not yet known. We therefore define the random variable $Y$ that represents the as-yet-unknown level of demand. The forecast $F$ specifies a predictive distribution for $Y$. Here we identify $F$ with its cumulative distribution function (CDF), and $F^{-1}$ denotes the quantile function. With this formalization of the decision-making task, we can proceed to develop a proper scoring rule using the procedure outlined in section \ref{sec:methods.decisiontheory}.

@@

We now develop a scoring rule for probabilistic forecasts that measures the value of a forecast as an input to decision making about how to allocate limited resources to meet demand across multiple locations. As a concrete example, we take the resource to be a good such as ventilators or oxygen supply. An administrator is tasked with determining where to send these resources so as to meet demand among hospital patients in different facilities or states. In contrast to the decision-making problem in the previous section, the administrator is not able to control the total amount of supply; rather, their task is to determine how to allocate the fixed supply to different locations.

In this decision-making setting, an action $\mathbf{x} = (x_1, \ldots, x_n)$ is a vector specifying the amount that is allocated to each of the $n$ locations. We require that each $x_i$ is non-negative and that the total allocation across all locations does not exceed a constraint $K$ on the total available resources: $\sum_{i=1}^n x_i \leq K$. \elr{Since we got rid of the C parameter, maybe we should just make this a hard constraint here, $\sum_i x_i = K$?} The set $\mathcal{X}$ collects all possible allocations that satisfy these constraints. The eventually realized resource demand in each location is denoted by $\mathbf{y} = (y_1, \ldots, y_n)$. Again, these levels of demand are not known at the time of decision making, so we define the random vector $Y = (Y_1, \ldots, Y_n)$ where $Y_i$ represents the as-yet-unknown level of resource demand in location $i$. Forecasts of demand in each location are collected in $F = (F_1, \ldots, F_n)$. We assume that the forecasts do not allow for the possibility of negative demand, i.e. the support of each $F_i$ is a subset of $\mathbb{R}^+$. As in the previous section, we assume that a loss $L$ is incurred for each unit of unmet need.

We note that a number of generalizations to this loss specification have been formulated in the literature, including an allowance for costs for over-allocation to a particular unit (e.g. if there are storage costs for unused resources), differing losses different units (e.g. if a unit of unmet demand imposes more severe costs in one location than another), and the introduction of a convex function that controls the rate at which costs accrue depending on the scale of need. We consider these and other generalizations in other work.

With this notation in place, we can develop a proper scoring rule following the outline in section \ref{sec:methods.decisiontheory}.

\paragraph{Step 1: specify a loss function.} The loss associated with a particular allocation is calculated by summing contributions from unmet demand in each location:
\begin{equation}
s_A(\mathbf{x}, \mathbf{y}; L) = \sum_{i=1}^n L (x_i - y_i)_-. \label{eqn:loss_fn}
\end{equation}
As in the previous section, $(x_i - y_i)_-$ is $0$ if the amount $x_i$ allocated to unit $i$ is greater than or equal to the realized demand $y_i$ in that location; otherwise, it is $y_i - x_i$, the amount of unmet need in that location.

\paragraph{Step 2: Given a probabilistic forecast $F$, identify the Bayes act.} The Bayes act is the allocation that minimizes the expected loss:
\begin{align}
    \mathbf{x}^F &= \underset{\mathbf{x}\in \mathbb{R}^N, 0 \leq \mathbf{x}}{\mathrm{arg min}}\,\, \sbar_A^F(\mathbf{x}; L) \text{ subject to } 
    \sum_{i=1}^N x_i \leq K \text{, where} \label{AP} \\
    \sbar_A^F(\mathbf{x}; L) &= \mathbb{E}_{F} [s_A(\mathbf{x}, \mathbf{Y}; L)] = \sum_{i=1}^{N} \mathbb{E}_{F_i}[s_A(x_i, Y_i; L)] \label{eqn:sbar}
\end{align}
It can be shown that with the loss function given in Equation \eqref{eqn:loss_fn}, the Bayes act has $x^F_i = F_i^{-1}(1 - \lambda^{\star}/L)$, where $\lambda^{\star}$ is chosen so as to satisfy the equation
\begin{align}
\sum_{i=1}^{N}F_i^{-1}(1 - \lambda^{\star}/L) = K.
\end{align}
This partial solution to the allocation problem seems to have first appeared in \cite{hadleywhitin1963}; see the supplemental materials for a derivation.

One interpretation of this result is that the Bayes act sets the allocation in each location $i$ to a quantile of the forecast distribution $F_i$ for that location. The quantile is at a probability level $\alpha = 1 - \lambda^{\star}/L$ that is the same for all locations, and is chosen such that the constraint is satisfied. An alternative interpretation comes from noting that for each location $i$, $\frac{\partial}{\partial x_i} \sbar_A^F(\mathbf{x}) \vert_{\mathbf{x}=\mathbf{x}^F} = \lambda^{\star}$ (see the supplement for a proof).
%$\frac{\partial}{\partial x_i} \sbar_F(\mathbf{x}) = L[F_i(x_i) - 1]$ (see the supplement for a derivation), and evaluating at the Bayes allocation we arrive at
%\begin{align}
%\frac{\partial}{\partial x_i} \sbar_F(\mathbf{x}) \vert_{\mathbf{x}=\mathbf{x}^F} &= L[F_i(x^F_i) - 1] = L[F_i(F_i^{-1}(1 - \lambda^{\star}/L)) - 1] = \lambda^{\star}.
%\end{align}
In words, at the allocation given by the Bayes act, the rate of change of the expected score as a function of the amount allocated to location $i$ is given by $\lambda^{\star}$.
This derivative is the same for all locations, so the optimal allocation divides the available resources across all locations in such a way that according to $F$, the expected benefit of 1 additional unit of resources is the same in all locations.
